Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-02-01 3.phase_transitions/3.global-tests.tex
merge of Ch. 3 and the new Ch. 4 & 5
I merged my work on Ch 3 and Ch. 4&5 with the "revision-GWAS" branch.

Need some more work -- putting the Mill's ratio in Ch 2 perhaps...

To the best of our knowledge, performance of simple statistics such as $L_1$, $L_2$ norms, and 
\stilian{the sum statistic $T$ in \eqref{eq:sum-statistic}}{R2, Typo 7}
in this weak signal setting have not been reported in the literature. %perhaps due to a perceived lack of novelty. 
Our first theorem establishes the performance of these simple but popular statistics for detecting sparse signals 
in high-dimensions, and summarizes the known results.

@@

{\em Part {\rm (iv)}:}  We now turn to the $L_2$-norm statistic.
Recall a non-central chi-square random variable $\chi^2_k(\lambda)$ has mean $(k+\lambda)$ and variance $2(k+2\lambda)$.
Since the observations have distributions $\mathrm{N}(0,1)$ under the null and $\mathrm{N}(p^r,1)$ under the alternative, we have $x^2(i)\sim \chi^2_1(0)$ for $i\not\in S$ and $x^2(i)\sim \chi^2_1(p^{2r})$ for $i\in S$.
Therefore, the mean and variance of the (centered and scaled) $L_2$ statistics are
\begin{equation}
    \E\left[\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right] = 
    \begin{cases}
    0 & \text{under } H_0\\
    p^{1-\beta}p^{2r}p^{-1/2} = p^{1/2-\beta+2r} & \text{under } H_1,
    \end{cases}
\end{equation}
and 
\begin{equation}\label{e:detection-variance}
    \mathrm{Var}\left(\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right) = 
    \begin{cases}
    \frac{1}{p}2p = 2 & \text{under } H_0\\
    \frac{1}{p}\left(2p+4p^{1-\beta+2r}\right) = 2(1+ 2p^{2r-\beta}) & \text{under } H_1,
    \end{cases}
\end{equation}
respectively.
By the central limit theorem, we have
\begin{equation}
    \frac{1}{\sqrt{2p}}\sum_{i=1}^p \left(x(i)^2-1\right) \implies \mathrm{N}(0,1),  
\end{equation}
under the null.    On the other hand, under the alternative, since $p^{2r-\beta}\to0$ for all $r<0$ and $\beta>0$, 
 the variance in \eqref{e:detection-variance} converges to $2$, as $p\to\infty$ and an application of the Lyapunov version of 
 the CLT, entails
 \begin{equation}
    \frac{1}{\sqrt{2p}}\left(\sum_{i=1}^p \left(x(i)^2-1\right) - p^{1/2-\beta+2r}\right) \implies \mathrm{N}(0,1).
\end{equation}
Hence, perfect detection with the $L_2$-norm is possible if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, if $r<\beta/2-1/4$, the distributions of the (scaled) statistics merge under the null and the alternative. % \fbox{Hellinger distance is scale-invariant (show), detection is impossible}.

>>>>

To the best of our knowledge, performance of simple statistics such as $L_1$, $L_2$ norms, and 
the sum statistic $T$ in \eqref{eq:sum-statistic} in this weak signal setting have not been reported in the literature. %perhaps due to a perceived lack of novelty. 
Our first theorem establishes the performance of these simple but popular statistics for detecting sparse signals 
in high-dimensions, and summarizes the known results.

@@

{\em Part {\rm (iv)}:}  We now turn to the $L_2$-norm statistic.
Recall a non-central chi-square random variable $\chi^2_k(\lambda)$ has mean $(k+\lambda)$ and variance $2(k+2\lambda)$.
Since the observations have distributions $\mathrm{N}(0,1)$ under the null and $\mathrm{N}(p^r,1)$ under the alternative, we have $x^2(i)\sim \chi^2_1(0)$ for $i\not\in S$ and $x^2(i)\sim \chi^2_1(p^{2r})$ for $i\in S$.
Therefore, the mean and variance of the (centered and scaled) $L_2$ statistics are
\begin{equation}
    \E\left[\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right] = 
    \begin{cases}
    0 & \text{under } H_0\\
    p^{1-\beta}p^{2r}p^{-1/2} = p^{1/2-\beta+2r} & \text{under } H_1,
    \end{cases}
\end{equation}
and 
\begin{equation}\label{e:detection-variance}
    \mathrm{Var}\left(\frac{1}{\sqrt{p}}\sum_{i=1}^p \left(x(i)^2-1\right)\right) = 
    \begin{cases}
    \frac{1}{p}2p = 2 & \text{under } H_0\\
    \frac{1}{p}\left(2p+4p^{1-\beta+2r}\right) = 2(1+ 2p^{2r-\beta}) & \text{under } H_1,
    \end{cases}
\end{equation}
respectively.
By the central limit theorem, we have
\begin{equation}
    {\color{red}\frac{1}{\sqrt{2p}}}\sum_{i=1}^p \left(x(i)^2-1\right) \implies \mathrm{N}(0,1),  
\end{equation}
under the null.    On the other hand, under the alternative, since $p^{2r-\beta}\to0$ for all $r<0$ and $\beta>0$, 
 the variance in \eqref{e:detection-variance} converges to $2$, as $p\to\infty$ and an application of the Lyapunov version of 
 the CLT, entails
 \begin{equation}
    {\color{red} \frac{1}{\sqrt{2p}}}\left(\sum_{i=1}^p \left(x(i)^2-1\right) - p^{1/2-\beta+2r}\right) \implies \mathrm{N}(0,1).
\end{equation}
Hence, perfect detection with the $L_2$-norm is possible if $p^{1/2-\beta+2r}$ diverges, i.e., $r>\beta/2-1/4$.
On the other hand, if $r<\beta/2-1/4$, the distributions of the (scaled) statistics merge under the null and the alternative. % \fbox{Hellinger distance is scale-invariant (show), detection is impossible}.

### 2021-02-01 3.phase_transitions/phase_transitions.tex
ch 3
almost done editing ch 3 (for formatting).  I need to absorb the auxuliary results from the appendix and check the proofs carefully once again. I like the new format where the proofs are presented right after the statments and the various support recovery criteria are presented in separate sections rather than sub-sections.  Thus, they are visible in the table of contents.
merge of Ch. 3 and the new Ch. 4 & 5
I merged my work on Ch 3 and Ch. 4&5 with the "revision-GWAS" branch.

Need some more work -- putting the Mill's ratio in Ch 2 perhaps...

\stilian{The purpose of this chapter is to provide a unified review of the fundamental statistical limits in the 
sparse signal detection and support recovery problems.  Our goal is to convey the main ideas and thus we shall focus on 
the simple but important setting of independent Gaussian errors.}{\xout{We study the fundamental limits of the signal detection and support recovery problems in the location models with independent Gaussian errors in this chapter.}}
Specifically, we derive the conditions under which the detection and support recovery problems succeed and fail in the sense of \eqref{eq:support-recovery-success} and \eqref{eq:support-recovery-failure}, in the additive error model 
\begin{equation} \label{eq:model-additive-Chapter3}
    x(i) = \mu(i) + \epsilon(i), \quad i=1,\ldots,p,
\end{equation}
where the errors $\epsilon(i)$'s are \ac{iid} standard Gaussians random variables.
Once again, we restrict our analysis to models with independent and identically distributed Gaussian errors for the 
moment. Both the distributional assumption and the independence assumption will be relaxed substantially in the following chapters.

@@

\section{Sparse signal support recovery problems}
\label{sec:additive-error-model-boundaries}
\input{3.phase_transitions/3.Gaussian-support-recovery.tex}

>>>>

The purpose of this chapter is to provide a unified review of the fundamental statistical limits in the 
sparse signal detection and support recovery problems.  Our goal is to convey the main ideas and thus we shall focus on 
the simple but important setting of independent Gaussian errors.
Specifically, we derive the conditions under which the detection and support recovery problems succeed and fail in the sense of \eqref{eq:support-recovery-success} and \eqref{eq:support-recovery-failure}, in the additive error model 
\begin{equation} \label{eq:model-additive-Chapter3}
    x(i) = \mu(i) + \epsilon(i), \quad i=1,\ldots,p,
\end{equation}
where the errors $\epsilon(i)$'s are \ac{iid} standard Gaussians random variables.
Once again, we restrict our analysis to models with independent and identically distributed Gaussian errors for the 
moment. Both the distributional assumption and the independence assumption will be relaxed substantially in the following chapters.

@@

%\newpage
\section{Sparse signal support recovery problems}
\label{sec:additive-error-model-boundaries}
\input{3.phase_transitions/3.Gaussian-support-recovery.tex}

### 2021-02-01 5.Bayes.optimality/5.Bayes_optimality_and_minimax.tex
merge of Ch. 3 and the new Ch. 4 & 5
I merged my work on Ch 3 and Ch. 4&5 with the "revision-GWAS" branch.

Need some more work -- putting the Mill's ratio in Ch 2 perhaps...

\section{A discussion on optimality}

@@

Consider the model \eqref{eq:model-additive} with independent errors. Theorem \ref{thm:necessary-strengthened} 
and Corollary \ref{cor:minimax-over-dependence} provide a nearly complete picture of the difficulty in the exact support 
recovery problem, in the regime when the thresholding estimators are optimal. Specifically, in such cases the signal classification boundary is universal.

 Proposition \ref{prop:likelihood-ratio-thresholding}, and indeed, Example \ref{exmp:suboptimal-data-thresholding} demonstrate that thresholding procedures are in fact \emph{sub-optimal} for $\text{AGG}(\nu)$ models with $\nu<1$.
Therefore, the optimality of thresholding procedures (specifically, Bonferroni's procedure) only applies to $\text{AGG}(\nu)$ models with $\nu\ge1$. 
% This suggests that the condition $\nu\ge1$ in Corollary \ref{cor:minimax-over-dependence} is tight.

If we restrict the space of methods to only thresholding procedures, then results in Section \ref{subsec:point-wise-minimax} state that the phase transition phenomenon --- the 0-1 law in the sense of Corollary \ref{cor:point-wise-minimax} --- is universal in all error models with rapidly varying tails.
This includes $\text{AGG}(\nu)$ models {\it for all} $\nu>0$.
In contrast, models with heavy (regularly varying) tailed errors do not exhibit this phenomenon (see Theorem \ref{thm:heavy-tails}).
We summarize the properties of thresholding procedures in Table \ref{table:role-of-thresholding}.

@@

\begin{table}[ht]
    \centering
    \caption{Properties of thresholding procedures under different error distributions when errors are independent. Properties of the error distributions are listed in brackets.}
    \medskip
    \begin{tabular}{p{40mm}p{37mm}p{37mm}} \toprule
        Thresholding procedure & Bayes optimality &  Phase transition \\ 
        (Error distributions) &  (Log-concave density) & (Rapidly-varying tails) \\ \midrule
        AGG($\nu$), $\nu\ge1$ & Yes (Yes) & Yes (Yes) \\ \cmidrule{1-3}
        AGG($\nu$), $0<\nu<1$ & No (No) & Yes (Yes) \\ \cmidrule{1-3}
        Power laws & No (No) & No (No) \\ \bottomrule
    \end{tabular}
    \label{table:role-of-thresholding}
\end{table}

>>>>

\section{Optimality and sub-optimality: A discussion}

@@

We conclude with a brief summary on the optimality and sub-optimality of the thresholding procedures in the problem of exact support estimation.
For clarity, we focus on the model \eqref{eq:model-additive} with {\em independent} errors.

Theorem \ref{thm:necessary-strengthened} and Corollary \ref{cor:minimax-over-dependence} provide a nearly complete picture of the difficulty 
in the exact support  recovery problem, in the regime when the thresholding estimators are optimal. Specifically, in such cases the signal classification boundary is universal.  On the other hand, Proposition \ref{prop:likelihood-ratio-thresholding}, and indeed, Example \ref{exmp:suboptimal-data-thresholding} demonstrate that thresholding procedures are \emph{sub-optimal} for $\text{AGG}(\nu)$ models with $\nu<1$. Therefore, the optimality of thresholding
procedures (specifically, Bonferroni's procedure) only applies to $\text{AGG}(\nu)$ models with $\nu\ge1$. 
% This suggests that the condition $\nu\ge1$ in Corollary \ref{cor:minimax-over-dependence} is tight.

If we restrict the space of methods to only thresholding procedures, then the results in Section \ref{subsec:point-wise-minimax} state that the phase 
transition phenomenon --- the 0-1 law in the sense of Corollary \ref{cor:point-wise-minimax} --- is universal in all error models with rapidly varying tails.
This includes $\text{AGG}(\nu)$ models {\it for all} $\nu>0$.  In contrast, models with heavy (regularly varying) tailed errors do not exhibit this phenomenon 
(form more details, see Theorem \ref{thm:heavy-tails}).  We summarize the properties of thresholding procedures in 
Table \ref{table:role-of-thresholding}.

@@

\begin{table}[ht]
    \centering
    \caption{Properties of thresholding procedures under different error distributions when the errors are independent. 
    Properties of the error distributions are listed in brackets.}
    \medskip
    \begin{tabular}{p{40mm}p{37mm}p{37mm}} \toprule
        Thresholding procedure & Bayes optimality &  Phase transition \\ 
        (Error distributions) &  (Log-concave density) & (Rapidly-varying tails) \\ \midrule
        AGG($\nu$), $\nu\ge1$ & Yes (Yes) & Yes (Yes) \\ \cmidrule{1-3}
        AGG($\nu$), $0<\nu<1$ & No (No) & Yes (Yes) \\ \cmidrule{1-3}
        Power laws & No (No) & No (No) \\ \bottomrule
    \end{tabular}
    \label{table:role-of-thresholding}
\end{table}
